---
title: "Getting Started with daggr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with daggr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 150
)
library(daggr)
```

## Why daggr?

Drawing DAGs with `ggdiagram` and `ggarrow` is powerful but requires ~30 lines
of boilerplate (helper functions, arrowhead config, theme setup). `daggr` wraps
all of that so you can focus on the graph itself.

## Canonical DAG Structures

### 1. Fork (Confounder)

A common cause **Z** creates a spurious association between **X** and **Y**.
Conditioning on **Z** blocks the backdoor path.

```{r fork, fig.cap = "Fork (Confounder)"}
Z <- dag_ellipse("Z")
X <- dag_ellipse("X") |> place(from = Z, where = "below left", sep = 2.5)
Y <- dag_ellipse("Y") |> place(from = Z, where = "below right", sep = 2.5)

ggdiagram(font_size = 14) +
  Z + X + Y +
  dag_connect(Z, X) +
  dag_connect(Z, Y) +
  theme_dag() +
  ggplot2::labs(title = "Fork (Confounder)")
```

### 2. Pipe (Mediator)

The effect of **X** on **Y** is transmitted entirely through the mediator **M**.
Conditioning on **M** blocks the causal path.

```{r pipe, fig.cap = "Pipe (Mediator)"}
X <- dag_ellipse("X")
M <- dag_ellipse("M") |> place(from = X, where = "right", sep = 2.5)
Y <- dag_ellipse("Y") |> place(from = M, where = "right", sep = 2.5)

ggdiagram(font_size = 14) +
  X + M + Y +
  dag_connect(X, M) +
  dag_connect(M, Y) +
  theme_dag() +
  ggplot2::labs(title = "Pipe (Mediator)")
```

### 3. Collider

**X** and **Y** are independent causes of **C**. Conditioning on the collider
**C** opens a spurious path between **X** and **Y**.

```{r collider, fig.cap = "Collider"}
X <- dag_ellipse("X")
Y <- dag_ellipse("Y") |> place(from = X, where = "right", sep = 5)
C <- dag_ellipse("C") |> place(from = X, where = "below right", sep = 2.5)

ggdiagram(font_size = 14) +
  X + Y + C +
  dag_connect(X, C) +
  dag_connect(Y, C) +
  theme_dag() +
  ggplot2::labs(title = "Collider")
```

### 4. Instrumental Variable

**Z** is an instrument: it affects **Y** only through **X**. The unmeasured
confounder **U** biases the naive **X -> Y** estimate, but the instrument
identifies the causal effect.

```{r iv, fig.cap = "Instrumental Variable"}
Z <- dag_ellipse("Z")
X <- dag_ellipse("X") |> place(from = Z, where = "right", sep = 2.5)
Y <- dag_ellipse("Y") |> place(from = X, where = "right", sep = 2.5)
U <- dag_ellipse("U", fill = "#708090") |>
  place(from = X, where = "above right", sep = 2.5)

ggdiagram(font_size = 14) +
  Z + X + Y + U +
  dag_connect(Z, X) +
  dag_connect(X, Y) +
  dag_connect(U, X) +
  dag_connect(U, Y) +
  theme_dag() +
  ggplot2::labs(title = "Instrumental Variable")
```

### 5. Front-Door (Smoking-Tar-Cancer)

Pearl's canonical example. Unmeasured **Genetics** confounds **Smoking** and
**Cancer**, but the front-door path through **Tar** allows identification.

```{r frontdoor, fig.cap = "Front-Door Criterion"}
Genetics <- dag_ellipse("Genetics", fill = "#708090")
Smoking  <- dag_ellipse("Smoking") |>
  place(from = Genetics, where = "below left", sep = 2.5)
Cancer   <- dag_ellipse("Cancer") |>
  place(from = Genetics, where = "below right", sep = 2.5)
Tar      <- dag_ellipse("Tar") |>
  place(from = Smoking, where = "below right", sep = 2.5)

ggdiagram(font_size = 14) +
  Genetics + Smoking + Cancer + Tar +
  dag_connect(Genetics, Smoking) +
  dag_connect(Genetics, Cancer) +
  dag_connect(Smoking, Tar) +
  dag_connect(Tar, Cancer) +
  theme_dag() +
  ggplot2::labs(title = "Front-Door Criterion")
```

### 6. M-Bias (Butterfly)

Adjusting for the collider **M** *increases* bias. **U1** and **U2** are
unmeasured, but conditioning on **M** opens a spurious path between **X** and
**Y**.

```{r mbias, fig.cap = "M-Bias (Butterfly)"}
X  <- dag_ellipse("X")
Y  <- dag_ellipse("Y")  |> place(from = X, where = "right", sep = 5)
U1 <- dag_ellipse("U1", fill = "#708090") |>
  place(from = X, where = "above right", sep = 2.5)
U2 <- dag_ellipse("U2", fill = "#708090") |>
  place(from = Y, where = "above left", sep = 2.5)
M  <- dag_ellipse("M") |>
  place(from = U1, where = "below right", sep = 2.5)

ggdiagram(font_size = 14) +
  X + Y + U1 + U2 + M +
  dag_connect(U1, X) +
  dag_connect(U1, M) +
  dag_connect(U2, M) +
  dag_connect(U2, Y) +
  dag_connect(X, Y) +
  theme_dag() +
  ggplot2::labs(title = "M-Bias (Butterfly)")
```

### 7. Simpson's Paradox (Berkeley Admissions)

**Gender** affects both **Department** choice and **Admission** directly. The
marginal association reverses when conditioning on **Department**.

```{r simpson, fig.cap = "Simpson's Paradox"}
Gender     <- dag_ellipse("Gender")
Department <- dag_ellipse("Department") |>
  place(from = Gender, where = "right", sep = 2.5)
Admission  <- dag_ellipse("Admission") |>
  place(from = Department, where = "right", sep = 2.5)

ggdiagram(font_size = 14) +
  Gender + Department + Admission +
  dag_connect(Gender, Department) +
  dag_connect(Department, Admission) +
  dag_connect(Gender, Admission, arc_bend = 0.4) +
  theme_dag() +
  ggplot2::labs(title = "Simpson's Paradox")
```

## Advanced Features

### DGP vs Measurement

Use `annotate()` to label layers, and `linetype` to distinguish causal
relationships from observed associations.

```{r dgp-measurement, fig.cap = "DGP vs Measurement", fig.width = 8, fig.height = 5}
# DGP layer
dgp_ideo  <- dag_ellipse("Idéologie")
dgp_behav <- dag_ellipse("Comportement") |>
  place(dgp_ideo, "right", sep = 4)

# Measurement layer
meas_ideo  <- dag_ellipse("Idéologie") |>
  place(dgp_ideo, "bottom", sep = 3)
meas_behav <- dag_ellipse("Comportement") |>
  place(meas_ideo, "right", sep = 4)

ggdiagram(font_size = 14) +
  dgp_ideo + dgp_behav +
  meas_ideo + meas_behav +
  dag_connect(dgp_ideo, dgp_behav, linetype = "solid") +
  dag_connect(meas_behav, meas_ideo, linetype = "dashed") +
  ggplot2::annotate("text",
    x = mean(c(dgp_ideo@center@x, dgp_behav@center@x)),
    y = 3, label = "DGP", fontface = "bold", size = 6) +
  ggplot2::annotate("text",
    x = mean(c(meas_ideo@center@x, meas_behav@center@x)),
    y = -3, label = "Measurement", fontface = "bold", size = 6) +
  theme_dag()
```

### Mixed Node Shapes and Linetypes

Use `dag_rectangle()` for observed/constructed variables and different
`linetype` values to encode edge semantics (e.g., definitional links vs causal
effects).

```{r polsci, fig.cap = "Political Attitudes DAG", fig.width = 9, fig.height = 7}
A <- dag_ellipse("Intérêt politique")
B <- dag_ellipse("Autoplacement<br>Gauche-Droite") |>
  place(A, "top left", sep = 3.5)
C <- dag_ellipse("A voté<br>à la dernière<br>élection") |>
  place(A, "top right", sep = 3.5)
D <- dag_rectangle("Extrémisme", fill = "darkred") |>
  place(B, "top", sep = 2.5)
E <- dag_rectangle("Participation<br>électorale", fill = "darkred") |>
  place(C, "top", sep = 2.5)

ggdiagram() +
  A + B + C + D + E +
  dag_connect(D, B, linetype = 4) +
  dag_connect(E, C, linetype = 4) +
  dag_connect(A, B) +
  dag_connect(A, C) +
  dag_connect(B, C) +
  dag_connect(D, E, linetype = 3) +
  theme_dag()
```

### Multiline Labels

Node labels support `<br>` for line breaks, which is useful for longer variable
names.

```{r multiline, fig.cap = "Multiline Labels"}
X <- dag_ellipse("Years of<br>Education")
M <- dag_ellipse("Political<br>Knowledge") |>
  place(X, "right", sep = 3.5)
Y <- dag_ellipse("Voter<br>Turnout") |>
  place(M, "right", sep = 3.5)

ggdiagram(font_size = 14) +
  X + M + Y +
  dag_connect(X, M) +
  dag_connect(M, Y) +
  dag_connect(X, Y, arc_bend = -0.4) +
  theme_dag()
```

### Custom Colors

Change `fill` to distinguish different types of variables (e.g., treatment,
outcome, unmeasured).

```{r colors, fig.cap = "Custom Node Colors", fig.width = 8, fig.height = 5}
Treatment <- dag_ellipse("Treatment", fill = "#2E7D32")
Outcome   <- dag_ellipse("Outcome", fill = "#C62828") |>
  place(Treatment, "right", sep = 4)
Confounder <- dag_ellipse("Confounder", fill = "#708090") |>
  place(Treatment, "above right", sep = 2.5)
Mediator   <- dag_ellipse("Mediator") |>
  place(Treatment, "below right", sep = 2.5)

ggdiagram(font_size = 14) +
  Treatment + Outcome + Confounder + Mediator +
  dag_connect(Treatment, Outcome) +
  dag_connect(Treatment, Mediator) +
  dag_connect(Mediator, Outcome) +
  dag_connect(Confounder, Treatment) +
  dag_connect(Confounder, Outcome) +
  theme_dag()
```
