---
title: "Getting Started with daggr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with daggr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  dpi = 150
)
library(daggr)
```

## Why daggr?

Drawing DAGs with `ggdiagram` and `ggarrow` is powerful but requires ~30 lines
of boilerplate (helper functions, arrowhead config, theme setup). `daggr` wraps
all of that so you can focus on the graph itself.

## 1. Fork (Confounder)

A common cause **Z** creates a spurious association between **X** and **Y**.
Conditioning on **Z** blocks the backdoor path.

```{r fork, fig.cap = "Fork (Confounder)"}
Z <- dag_ellipse("Z")
X <- dag_ellipse("X") |> place(from = Z, where = "below left", sep = 2.5)
Y <- dag_ellipse("Y") |> place(from = Z, where = "below right", sep = 2.5)

ggdiagram(font_size = 14) +
  Z + X + Y +
  dag_connect(Z, X) +
  dag_connect(Z, Y) +
  theme_dag() +
  ggplot2::labs(title = "Fork (Confounder)")
```

## 2. Pipe (Mediator)

The effect of **X** on **Y** is transmitted entirely through the mediator **M**.
Conditioning on **M** blocks the causal path.

```{r pipe, fig.cap = "Pipe (Mediator)"}
X <- dag_ellipse("X")
M <- dag_ellipse("M") |> place(from = X, where = "right", sep = 2.5)
Y <- dag_ellipse("Y") |> place(from = M, where = "right", sep = 2.5)

ggdiagram(font_size = 14) +
  X + M + Y +
  dag_connect(X, M) +
  dag_connect(M, Y) +
  theme_dag() +
  ggplot2::labs(title = "Pipe (Mediator)")
```

## 3. Collider

**X** and **Y** are independent causes of **C**. Conditioning on the collider
**C** opens a spurious path between **X** and **Y**.

```{r collider, fig.cap = "Collider"}
X <- dag_ellipse("X")
Y <- dag_ellipse("Y") |> place(from = X, where = "right", sep = 5)
C <- dag_ellipse("C") |> place(from = X, where = "below right", sep = 2.5)

ggdiagram(font_size = 14) +
  X + Y + C +
  dag_connect(X, C) +
  dag_connect(Y, C) +
  theme_dag() +
  ggplot2::labs(title = "Collider")
```

## 4. Instrumental Variable

**Z** is an instrument: it affects **Y** only through **X**. The unmeasured
confounder **U** biases the naive **X -> Y** estimate, but the instrument
identifies the causal effect.

```{r iv, fig.cap = "Instrumental Variable"}
Z <- dag_ellipse("Z")
X <- dag_ellipse("X") |> place(from = Z, where = "right", sep = 2.5)
Y <- dag_ellipse("Y") |> place(from = X, where = "right", sep = 2.5)
U <- dag_ellipse("U", fill = "#708090") |>
  place(from = X, where = "above right", sep = 2.5)

ggdiagram(font_size = 14) +
  Z + X + Y + U +
  dag_connect(Z, X) +
  dag_connect(X, Y) +
  dag_connect(U, X) +
  dag_connect(U, Y) +
  theme_dag() +
  ggplot2::labs(title = "Instrumental Variable")
```

## 5. Front-Door (Smoking-Tar-Cancer)

Pearl's canonical example. Unmeasured **Genetics** confounds **Smoking** and
**Cancer**, but the front-door path through **Tar** allows identification.

```{r frontdoor, fig.cap = "Front-Door Criterion"}
Genetics <- dag_ellipse("Genetics", fill = "#708090")
Smoking  <- dag_ellipse("Smoking") |>
  place(from = Genetics, where = "below left", sep = 2.5)
Cancer   <- dag_ellipse("Cancer") |>
  place(from = Genetics, where = "below right", sep = 2.5)
Tar      <- dag_ellipse("Tar") |>
  place(from = Smoking, where = "below right", sep = 2.5)

ggdiagram(font_size = 14) +
  Genetics + Smoking + Cancer + Tar +
  dag_connect(Genetics, Smoking) +
  dag_connect(Genetics, Cancer) +
  dag_connect(Smoking, Tar) +
  dag_connect(Tar, Cancer) +
  theme_dag() +
  ggplot2::labs(title = "Front-Door Criterion")
```

## 6. M-Bias (Butterfly)

Adjusting for the collider **M** *increases* bias. **U1** and **U2** are
unmeasured, but conditioning on **M** opens a spurious path between **X** and
**Y**.

```{r mbias, fig.cap = "M-Bias (Butterfly)"}
X  <- dag_ellipse("X")
Y  <- dag_ellipse("Y")  |> place(from = X, where = "right", sep = 5)
U1 <- dag_ellipse("U1", fill = "#708090") |>
  place(from = X, where = "above right", sep = 2.5)
U2 <- dag_ellipse("U2", fill = "#708090") |>
  place(from = Y, where = "above left", sep = 2.5)
M  <- dag_ellipse("M") |>
  place(from = U1, where = "below right", sep = 2.5)

ggdiagram(font_size = 14) +
  X + Y + U1 + U2 + M +
  dag_connect(U1, X) +
  dag_connect(U1, M) +
  dag_connect(U2, M) +
  dag_connect(U2, Y) +
  dag_connect(X, Y) +
  theme_dag() +
  ggplot2::labs(title = "M-Bias (Butterfly)")
```

## 7. Simpson's Paradox (Berkeley Admissions)

**Gender** affects both **Department** choice and **Admission** directly. The
marginal association reverses when conditioning on **Department**.

```{r simpson, fig.cap = "Simpson's Paradox"}
Gender     <- dag_ellipse("Gender")
Department <- dag_ellipse("Department") |>
  place(from = Gender, where = "right", sep = 2.5)
Admission  <- dag_ellipse("Admission") |>
  place(from = Department, where = "right", sep = 2.5)

ggdiagram(font_size = 14) +
  Gender + Department + Admission +
  dag_connect(Gender, Department) +
  dag_connect(Department, Admission) +
  dag_connect(Gender, Admission, arc_bend = 0.4) +
  theme_dag() +
  ggplot2::labs(title = "Simpson's Paradox")
```
